Modelling M/EEG data with Bayesian multilevel generalised additive models

### 2025-02-10 brms_meeg.qmd
reshaping the document and drafting a script to conduct proper simulations of onset/offset recovery

---
title: "Modelling M/EEG data with Bayesian nonparametric multilevel models"
shorttitle: "Bayesian M/EEG modelling"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive process and how they unfold over time. Typically, we are interested in testing whether such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset and offset of effects (e.g., differences across conditions). Here, we introduce a novel *model-based approch* for analysing one-dimensional M/EEG timeseries such as ERPs or decoding timecourses and their differences across conditions. This approach relies on Bayesian nonparametric multilevel modelling (multilevel generalised additive models or Gaussian processes), which outputs posterior probabililities of the effect being above chance at every timestep/voxel, while naturally taking into account the temporal dependencies and between-subject variability present in such data."
keywords: [EEG, MEG, generalised additive models, gaussian processes, mixed-effects models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

@@

@@

Recent example of GLM for EEG [@wüllhorst2025; @fischer2013]... "Similar approaches have been successfully applied to EEG time- (Rousselet et al., 2008) and frequency-domain (Cohen and Cavanagh, 2011) data and allow the simultaneous investigation of multiple independent variables while preserving the high temporal resolution of the EEG."... See also [@rousselet2008; @hauk2006]... Example of two-stage regression analysis [individual-level then group-level, 10.1101/2024.04.10.588861, @dunagan2024]...

@@

# defining simulation parameters
n_trials <- 50 # number of trials
n_ppt <- 20 # number of participants
outvar <- 1 # noise variance
srate <- 500 # sampling rate in Hz
ronset <- seq(150, 170, 2) # random onset for each participant

@@

# defining the true onset and offset in seconds
true_onset <- 0.160
true_offset <- 0.342

# plotting the data
raw_df %>%
    summarise(eeg = mean(eeg), .by = c(participant, condition, time) ) %>%
    ggplot(aes(x = time, y = eeg, colour = condition) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~participant) +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")
```

@@

```{r fig-erp, echo = FALSE, out.width = "75%", fig.cap = "Group-level average difference between conditions (mean +/- standard error of the mean). The 'true' (population value of the) onset is indicated by the vertical dashed line."}
# averaging across participants
raw_df %>%
    pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
    mutate(eeg_diff = cond2 - cond1) %>%
    summarise(
        eeg_mean = mean(eeg_diff),
        eeg_sem = sd(eeg_diff) / sqrt(n() ),
        .by = time
        ) %>%
    mutate(lower = eeg_mean - eeg_sem, upper = eeg_mean + eeg_sem) %>%
    # plotting the data
    ggplot(aes(x = time, y = eeg_mean) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_ribbon(
        aes(ymin = lower, ymax = upper, colour = NULL),
        alpha = 0.25, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    labs(x = "Time (s)", y = "EEG difference (a.u.)")
```

@@

We then fitted a GAM and GP regression model using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]...

@@

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "cr", k = 10, by = condition),
    data = ppt_df,
    family = gaussian(),
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/gam.rds"
    )
```

Now we fit the GP...

@@

    post_preds %>%
        ggplot(aes(x = time, y = eeg, colour = condition, fill = condition) ) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_point(
            data = data %>% summarise(eeg = mean(eeg), .by = c(time, condition) ),
            aes(fill = NULL),
            pch = 21, show.legend = FALSE
            ) +
        geom_line(aes(y = estimate__), show.legend = FALSE) +
        geom_ribbon(
            aes(ymin = lower__, ymax = upper__, colour = NULL),
            alpha = 0.25, show.legend = FALSE
            ) +
        labs(x = "Time (s)", y = "EEG signal (a.u.)")

}
```

```{r gam-gp-preds, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior predictions of the GAM (left) and GP (right) models."}
# plotting the posterior predictions
plot_post_preds(model = gam) + plot_post_preds(model = gp_model)
```

@@

We can retrieve the posterior probability of the slope for `condition` at each timestep.

```{r, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Slope for the difference between conditions according to the GAM (left) or the GP (right)."}
# check posterior_smooths()?
# https://www.rdocumentation.org/packages/brms/versions/2.22.0/topics/posterior_smooths.brmsfit
# defining a function to plot posterior slope
plot_post_slope <- function (model, data = NULL, eps = 0.1) {

@@

    # plotting it
    prob_y_above_0 %>%
        ggplot(aes(x = time, y = m) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_line() +
        geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
        labs(x = "Time (s)", y = expression(Pr(beta~"|"~data, model) ) )

@@

We can also compute the posterior probability of the slope for `condition` being above 0 (@fig-post-prob-test).

```{r fig-post-prob-test, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior probability of the EEG signal being above 0+eps according to the GAM (left) and the GP (right)."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data = NULL, eps = 0.1) {

@@

@@

    # computing the probability that y is above 0
    prob_y_above_0 <- data.frame(
        time = time_seq$time,
        m = colMeans(posterior_samples > (0 + eps) ),
        lower = apply(
            X = posterior_samples > (0 + eps),
            MARGIN = 2, FUN = quantile, probs = 0.025
            ),
        upper = apply(
            X = posterior_samples > (0 + eps),
            MARGIN = 2, FUN = quantile, probs = 0.975
            )
        )

    # plotting it
    prob_y_above_0 %>%
    ggplot(aes(x = time, y = m) ) +
    geom_hline(yintercept = 0.5, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line() +
    labs(x = "Time (s)", y = expression(Pr(beta>0~"|"~data, model) ) )

@@

We can also express this as the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional BF thresholds (@fig-post-prob-ratio). Also check this paper (<https://link.springer.com/article/10.3758/s13428-021-01613-6?fromPaywallRec=false>)?

```{r fig-post-prob-ratio, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Ratio of posterior probability according to the GAM (left) and the GP (right). Timesteps above threshold (>10) are highlighted in green."}
# defining a sequence of x values to make predictions
n_approx <- 1000
time_seq <- crossing(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx),
    condition = c("cond1", "cond2")
    ) %>%
    arrange(condition)

@@

# computing the probability that y is above 0
eps <- 0.05
prob_y_above_0 <- data.frame(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx)
    ) %>%
    mutate(post_prob = colMeans(posterior_samples) ) %>%
    mutate(m = colMeans(posterior_samples > eps) ) %>%
    mutate(prob_ratio = m / (1 - m) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms.
threshold <- 10
exceeding_times_gam <- prob_y_above_0 %>%
    dplyr::filter(prob_ratio > threshold) %>%
    summarise(cluster_onset = min(time), cluster_offset = max(time) ) %>%
    mutate(
        cluster_peak = prob_y_above_0 %>%
            dplyr::filter(post_prob == max(post_prob) ) %>%
            pull(time)
        ) %>%
    mutate(true_onset = 0.160, true_peak = 0.250, true_offset = 0.342) %>%
    mutate(model = "GAM")

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = time, y = prob_ratio) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line() +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 15,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
    labs(
        x = "Time (s)",
        y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) )
        )

# retrieving posterior samples
posterior_samples <- posterior_epred(object = gp_model, newdata = time_seq)

# computing the probability that y is above 0
prob_y_above_0 <- data.frame(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx)
    ) %>%
    mutate(post_prob = colMeans(posterior_samples) ) %>%
    mutate(m = colMeans(posterior_samples > eps) ) %>%
    mutate(prob_ratio = m / (1 - m) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms
exceeding_times_gp <- prob_y_above_0 %>%
    dplyr::filter(prob_ratio > threshold) %>%
    summarise(cluster_onset = min(time), cluster_offset = max(time) ) %>%
    mutate(
        cluster_peak = prob_y_above_0 %>%
            dplyr::filter(post_prob == max(post_prob) ) %>%
            pull(time)
        ) %>%
    mutate(true_onset = 0.160, true_peak = 0.250, true_offset = 0.342) %>%
    mutate(model = "GP")

# plotting it
gp_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = time, y = prob_ratio) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line() +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 15,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
    labs(
        x = "Time (s)",
        y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) )
        )

# plotting results for the two models
gam_ratio_plot + gp_ratio_plot

@@

@@

Below we are plotting the error function (where the error is computed as $\text{error}:=|\hat{\theta}-\theta|$) for both the onset and offset values of the ERP difference, according to various `eps` and `threshold` values. Remember that the signal is generated from a truncated Gaussian defining an objective onset at 160 ms, a maximum at 250 ms, and an offset at 342 ms. @fig-onset-error shows that the GP model can *exactly* recover the true onset and offset values, given some reasonable choice of `eps` and `threshold` values.

```{r fig-onset-error, echo = FALSE, dpi = 300, fig.width = 8, fig.asp = 0.5, fig.cap = "Error function of onset (left) and offset (right) estimation according to various eps and threshold values (according to the GP model). Minimum error values are indicated by red crosses."}
# defining values of eps and threshold to test
eps_values <- seq(from = 0, to = 0.2, by = 0.01)
threshold_values <- seq(from = 1, to = 50, by = 1)

# using as many values of time as in the original data
n_approx <- n_distinct(ppt_df$time)

>>>>

---
title: "Modelling M/EEG data with Bayesian multilevel generalised additive models"
shorttitle: "Bayesian M/EEG modelling"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive process and how they unfold over time. Typically, we are interested in testing whether such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics through followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset and offset of effects (e.g., differences across conditions). Here, we introduce a novel *model-based approch* for analysing one-dimensional M/EEG timeseries such as ERPs or decoding timecourses and their differences across conditions or group. This approach relies on Bayesian nonparametric multilevel modelling (multilevel generalised additive models), which outputs the posterior probabilility of the effect being above 0 (or above chance) at every timestep, while naturally taking into account the temporal dependencies and between-subject variability present in such data."
keywords: [EEG, MEG, generalised additive models, gaussian processes, mixed-effects models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

@@

From @dimigen2021: "Recently, spline regression has been applied to ERPs (Hendrix, Baayen, & Bolger, 2017; Kryuchkova,Tucker,Wurm,&Baayen,2012;Tremblay& Baayen, 2010; Tremblay & Newman, 2015)"...

@@

Recent example of GLM for EEG [@wüllhorst2025; @fischer2013]... "Similar approaches have been successfully applied to EEG time- (Rousselet et al., 2008) and frequency-domain (Cohen and Cavanagh, 2011) data and allow the simultaneous investigation of multiple independent variables while preserving the high temporal resolution of the EEG."... See also [@rousselet2008; @hauk2006]... Example of two-stage regression analysis [i.e., individual-level then group-level, @dunagan2024]...

@@

# defining simulation parameters
n_trials <- 50 # number of trials
n_ppt <- 20 # number of participants
outvar <- 1 # noise variance
srate <- 500 # sampling rate in Hz
ronset <- seq(from = 150, to = 170, by = 2) # random onset for each participant

@@

# defining the true onset and offset in seconds
true_onset <- 0.160
true_peak <- 0.250
true_offset <- 0.342

# plotting the data
raw_df %>%
    summarise(eeg = mean(eeg), .by = c(participant, condition, time) ) %>%
    ggplot(aes(x = time, y = eeg, colour = condition) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~participant) +
    scale_colour_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")
```

@@

```{r fig-erp, echo = FALSE, out.width = "75%", fig.cap = "Group-level average difference between conditions (mean +/- standard error of the mean). The 'true' onset and offset are indicated by the vertical dashed lines."}
# averaging across participants
raw_df %>%
    pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
    mutate(eeg_diff = cond2 - cond1) %>%
    summarise(
        eeg_mean = mean(eeg_diff),
        eeg_sem = sd(eeg_diff) / sqrt(n() ),
        .by = time
        ) %>%
    mutate(lower = eeg_mean - eeg_sem, upper = eeg_mean + eeg_sem) %>%
    # plotting the data
    ggplot(aes(x = time, y = eeg_mean) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_ribbon(
        aes(ymin = lower, ymax = upper, colour = NULL),
        alpha = 0.25, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    labs(x = "Time (s)", y = "EEG difference (a.u.)")
```

@@

We then fitted a GAM and GP regression model using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms`, that is, weakly informative priors. Four Markov Chain Monte-Carlo (MCMC) were ran for each model to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statisti $\hat{R}$.

@@

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "cr", k = 10, by = condition),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/gam.rds"
    )
```

Now we fit the GP regression model...

@@

    post_preds %>%
        ggplot(aes(x = time, y = eeg, colour = condition, fill = condition) ) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_vline(xintercept = true_offset, linetype = 2) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_point(
            data = data %>% summarise(eeg = mean(eeg), .by = c(time, condition) ),
            aes(fill = NULL),
            pch = 21, show.legend = FALSE
            ) +
        geom_line(aes(y = estimate__), show.legend = FALSE) +
        geom_ribbon(
            aes(ymin = lower__, ymax = upper__, colour = NULL),
            alpha = 0.25, show.legend = FALSE
            ) +
        scale_colour_met_d(name = "Johnson") +
        scale_fill_met_d(name = "Johnson") +
        labs(x = "Time (s)", y = "EEG signal (a.u.)")

```

```{r gam-gp-preds, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior predictions of the GAM (left) and GP (right) models."}
# plotting the posterior predictions
plot_post_preds(model = gam) + plot_post_preds(model = gp_model)

```

@@

We can retrieve the posterior probability of the slope for `condition` at each timestep (@fig-plot-post-slope).

```{r fig-plot-post-slope, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Slope for the difference between conditions according to the GAM (left) or the GP (right)."}
# defining a function to plot posterior slope
plot_post_slope <- function (model, data = NULL, eps = 0.1) {

@@

    # plotting it
    prob_y_above_0 %>%
        ggplot(aes(x = time, y = m) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_line() +
        geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
        scale_colour_met_d(name = "Johnson") +
        scale_fill_met_d(name = "Johnson") +
        labs(x = "Time (s)", y = expression(Pr(beta~"|"~data, model) ) )

@@

We can also compute the posterior probability of the slope for `condition` being above $0+\epsilon$ (@fig-post-prob-test), with $\epsilon$ defined as 10% of the standard deviation of the raw EEG signal.

```{r fig-post-prob-test, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior probability of the ERP difference (slope) being above 0 according to the GAM (left) and the GP (right)."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data = NULL, eps = 0.1, n_approx = 1e3) {

@@

    # defining a sequence of time values to make predictions
    time_seq <- crossing(
        time = seq(min(data$time), max(data$time), length.out = n_approx),
        condition = c("cond1", "cond2")
        ) %>%
        arrange(condition)

@@

    # computing the difference between cond2 and cond1
    posterior_samples <- posterior_samples[, (nrow(time_seq)/2+1):nrow(time_seq)] -
        posterior_samples[, 1:(nrow(time_seq)/2)]

    # computing the probability that y is above 0
    prob_y_above_0 <- data.frame(
        time = time_seq$time,
        m = colMeans(posterior_samples > (0 + eps) )
        # lower = apply(
        #     X = posterior_samples > (0 + eps),
        #     MARGIN = 2, FUN = quantile, probs = 0.025
        #     ),
        # upper = apply(
        #     X = posterior_samples > (0 + eps),
        #     MARGIN = 2, FUN = quantile, probs = 0.975
        #     )
        # )
        )

    # plotting it
    prob_y_above_0 %>%
        ggplot(aes(x = time, y = m) ) +
        geom_hline(yintercept = 0.5, linetype = 2) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_vline(xintercept = true_offset, linetype = 2) +
        geom_line() +
        ylim(c(0, 1) ) +
        labs(x = "Time (s)", y = expression(Pr(beta>0~"|"~data, model) ) )

@@

We can also express this as the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional thresholds on evidence ratios (@fig-post-prob-ratio).

```{r fig-post-prob-ratio, echo = FALSE, fig.width = 8, fig.asp = 0.75, fig.cap = "Ratio of posterior probability according to the GAM. Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.1 * sd(ppt_df$eeg)
# eps <- 0.1
threshold <- 10
true_peak <- 0.250

# defining a sequence of x values to make predictions
time_seq <- crossing(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx),
    condition = c("cond1", "cond2")
    ) %>%
    arrange(condition)

@@

# computing the probability that y is above 0
prob_y_above_0 <- data.frame(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx)
    ) %>%
    mutate(post_prob = colMeans(posterior_samples) ) %>%
    mutate(m = colMeans(posterior_samples > eps) ) %>%
    mutate(prob_ratio = m / (1 - m) ) %>%
    mutate(prob_ratio = ifelse(is.infinite(prob_ratio), 1e4, prob_ratio) ) %>%
    mutate(prob_ratio = ifelse(prob_ratio == 0, 0.0001, prob_ratio) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms.
exceeding_times_gam <- prob_y_above_0 %>%
    dplyr::filter(prob_ratio > threshold) %>%
    summarise(cluster_onset = min(time), cluster_offset = max(time) ) %>%
    mutate(
        cluster_peak = prob_y_above_0 %>%
            dplyr::filter(post_prob == max(post_prob) ) %>%
            pull(time)
        ) %>%
    mutate(true_onset = true_onset, true_peak = true_peak, true_offset = true_offset) %>%
    mutate(model = "GAM")

# creates grid for background gradient
# Create a grid covering the range of time and log-transformed y-values
bg_data <- expand.grid(
    time = seq(min(prob_y_above_0$time), max(prob_y_above_0$time), length.out = 100),
    # prob_ratio = 10^seq(log10(0.0001), log10(1e4), length.out = 100)
    prob_ratio = 10^seq(log10(min(prob_y_above_0$prob_ratio) ), log10(max(prob_y_above_0$prob_ratio) ), length.out = 100)
    )

# assigning a fill value, for example, using prob_ratio or another variable
bg_data$fill_value <- log10(bg_data$prob_ratio)

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
# prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = time, y = prob_ratio) ) +
    geom_tile(
        data = bg_data,
        aes(x = time, y = prob_ratio, fill = fill_value),
        alpha = 0.4,
        inherit.aes = FALSE,
        show.legend = FALSE
        ) +
    scale_fill_scico(palette = "vik", midpoint = log10(1) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line(linewidth = 0.8) +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 16,
        size = 3,
        na.rm = TRUE
        ) +
    # scale_y_log10(labels = label_log(digits = 2), limits = c(0.0001, 1e4) ) +
    scale_y_log10(
        labels = label_log(digits = 2),
        breaks = c(1/1e4, 1/1e3, 1/1e2, 1/10, 1, 10, 1e2, 1e3, 1e4),
        limits = c(0.0001, 1e4)
        ) +
    coord_cartesian(expand = FALSE) +
    labs(
        x = "Time (s)",
        y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) )
        )

# retrieving posterior samples
# posterior_samples <- posterior_epred(object = gp_model, newdata = time_seq)

# computing the difference between cond2 and cond1
# posterior_samples <- posterior_samples[, (ncol(posterior_samples)/2+1):ncol(posterior_samples)] -
#     posterior_samples[, 1:(ncol(posterior_samples)/2)]

# computing the probability that y is above 0
# prob_y_above_0 <- data.frame(
#     time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx)
#     ) %>%
#     mutate(post_prob = colMeans(posterior_samples) ) %>%
#     mutate(m = colMeans(posterior_samples > eps) ) %>%
#     mutate(prob_ratio = m / (1 - m) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms
# exceeding_times_gp <- prob_y_above_0 %>%
#     dplyr::filter(prob_ratio > threshold) %>%
#     summarise(cluster_onset = min(time), cluster_offset = max(time) ) %>%
#     mutate(
#         cluster_peak = prob_y_above_0 %>%
#             dplyr::filter(post_prob == max(post_prob) ) %>%
#             pull(time)
#         ) %>%
#     mutate(true_onset = true_onset, true_peak = true_peak, true_offset = true_offset) %>%
#     mutate(model = "GP")

# plotting it
# gp_ratio_plot <- prob_y_above_0 %>%
#     mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
#     ggplot(aes(x = time, y = prob_ratio) ) +
#     geom_hline(yintercept = 1, linetype = "dashed") +
#     geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
#     geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
#     geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
#     geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
#     geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
#     geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
#     geom_vline(xintercept = true_onset, linetype = 2) +
#     geom_vline(xintercept = true_offset, linetype = 2) +
#     geom_line() +
#     geom_point(
#         data = . %>% dplyr::filter(!is.na(above_thres) ),
#         aes(y = threshold),
#         colour = "darkgreen",
#         shape = 15,
#         size = 3,
#         na.rm = TRUE
#         ) +
#     scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
#     labs(
#         x = "Time (s)",
#         y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) )
#         )

# plotting results for the two models
# gam_ratio_plot + gp_ratio_plot
gam_ratio_plot

@@

## Multilevel modelling using ERP summary statistics

Next we fit a hierarchical/multilevel GAM using summary statistics of ERPs (mean and SD) at the participant level (similar to what is done in meta-analysis).

# fitting the GAM
meta_gam <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 10, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/meta_gam.rds"
    )
```

```{r meta-gp-preds, fig.width = 9, fig.asp = 0.5, fig.cap = "Posterior predictions for the hierarchical GAM and GP."}
# plotting the posterior predictions
meta_gam_preds <- plot(
    conditional_effects(x = meta_gam, effect = "time:condition"),
    points = FALSE, theme = theme_light(), plot = FALSE
    )[[1]] +
    scale_colour_met_d(name = "Johnson") +
    scale_fill_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")
meta_gp_preds <- plot(
    conditional_effects(x = meta_gp, effect = "time:condition"),
    points = FALSE, theme = theme_light(), plot = FALSE
    )[[1]] +
    scale_colour_met_d(name = "Johnson") +
    scale_fill_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")

@@

We then computed the difference between the true and estimated onset/offset of the ERP difference ($\text{error}:=|\hat{\theta}-\theta|$), according to various `eps` and `threshold` values. Remember that the signal is generated from a truncated Gaussian with an objective onset at 160 ms, a maximum at 250 ms, and an offset at 342 ms. @fig-onset-error shows that the hierarchical GAM can *exactly* recover the true onset and offset values, given some reasonable choice of `eps` and `threshold` values (e.g., a threshold of 20).

```{r fig-onset-error, echo = FALSE, dpi = 300, fig.width = 8, fig.asp = 0.5, fig.cap = "Error function of onset (left) and offset (right) estimation according to various eps and threshold values (according to the hierarchical GAM). Minimum error values are indicated by red crosses."}
# defining values of eps and threshold to test
eps_values <- seq(from = 0, to = 0.1, by = 0.01)
threshold_values <- seq(from = 1, to = 30, by = 1)

# using as many values of time as in the original data
# n_approx <- n_distinct(ppt_df$time)
